---
title: "Semi-parametric Benchmark Dosing with semibmd"
author: "Alex Stringer"
date: "`r Sys.Date()`"
output: github_document
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

# Install the `semibmd` package

The package is hosted at https://github.com/awstringer1/semibmd. It's currently a private repo so you can't install using e.g. `remotes::install_github`. Instead, clone the repository to a local directory, say
`~/work/projects/benchmark-dose/code/semibmd` for example,
and then in `R` do:
```{r install_it}
install.packages(pkgs='~/work/projects/benchmark-dose/code/semibmd',repos=NULL,type='source')
library(semibmd)
```

# Simulate some data

Simulate some data to which the dose-response model is to be fit:
```{r simulate_data}
n <- 1000
f <- function(x) exp(-5*x)
xmin <- 0
xmax <- 1
xcov <- seq(xmin,xmax,length.out=n)
x1 <- runif(n,xmin,xmax)
x2 <- runif(n,xmin,xmax)
set.seed(43798)
dat <- data.frame(y = rnorm(n,f(xcov)+2*x1-x2,1),x = xcov,x1=x1,x2=x2)
head(dat)
```

The exposure variable is `x` and there are two additional variables `x1` and `x2` that
can be included in the model.

# Benchmark dosing

## Fit the model

Fit the dose-response model and calculate the benchmark dose information:
```{r fitmodel}
mod <- benchmark_dose(y~s(x,bs='mpd')+x1+x2,
                      data=dat,
                      exposure = 'x',
                      x0=0,
                      p0=.05,
                      BMR=.05,
                      monotone = TRUE)
```

The first argument to `benchmark_dose` is a formula compatible with `scam::scam` (if `monotone=TRUE`) or with `mgcv::gam` (if `monotone=FALSE`). The `scam` package accepts
formulas just like `gam`, with the addition that terms you want to be monotone should
have their bases specified manually to be one of the special monotone spline basis. See
`?scam::scam` for a list. Here I chose `bs='mpd'` for "monotone p-spline, decreasing", to
enforce a monotone-decreasing estimated curve.

If you choose `monotone = TRUE` then `scam` is used; I added the option to choose `monotone=FALSE` and fit an ordinary `gam`, based on our discussions. However, I have not yet tested this feature.

The rest of the arguments are fairly self-explanatory. You have to tell it which
variable is the exposure variable (that you want the BMD(L) for) using `exposure=...`. You
can look at `?benchmark_dose` for the full documentation.

## Get the summaries

Summaries and plots are obtained the usual way:
```{r summary_plot}
summary(mod)
plot(mod)
```

The `summary` method just appends the estimated BMD(L) onto the summary from the
dose-response model. The `plot` method just adds vertical lines to the plotted `scam/gam`
at the BMD and BMDL.

For more detailed access, you can get the actual fitted model using `get_model(mod)`
and the estimated BMD(L) using `get_bmd(mod)`.

## Regular non-monotone GAM

We can also fit a regular `gam` by setting `montone = FALSE`. In this case, you change the basis in the formula to be
something compatible with `mgcv`:
```{r fitmodelgam}
mod <- benchmark_dose(y~s(x,bs='bs')+x1+x2, # B-spline
                      data=dat,
                      exposure = 'x',
                      x0=0,
                      p0=.05,
                      BMR=.05,
                      monotone = FALSE)
```

Summaries and plots are obtained the same way:
```{r summary_plot_gam}
summary(mod)
plot(mod)
```

# Alternative BMDL Calculation

I have added functionality for computing BMDLs using the delta method and bootstrapping. The argument `BMDL` in the
call to `benchmark_dose` controls which BMDL(s) is/are calculated; current options are `all` (default), `score` and `delta`.
Note that the contents of the `bmdl` slot in the output object is never anything other than the `score` BMDL; the `delta` BMDL
is stored elsewhere and accessed using a getter function, see below.

The argument `boot` controls the number of (parametric) bootstrap iterations, with default `0` meaning don't bootstrap. Like the `delta` BMDL,
a bootstrapped BMDL will be stored elsewhere in the output object and accessed using a getter.

Here is how to fit it:
```{r fitmodelallbmdl}
mod <- benchmark_dose(y~s(x,bs='bs')+x1+x2,
                      data=dat,
                      exposure = 'x',
                      x0=0,
                      p0=.05,
                      BMR=.05,
                      monotone = TRUE,
                      BMDL = 'all',
                      boot = 200
)
```

The `summary` method still returns just the `score` BMDL. Here are the various getter functions to get everything:
```{r getters}
# BMD and score BMDL:
get_bmd(mod)
# All BMDLs:
get_all_bmdl(mod)
```

# Errors and diagnostics

The `benchmark_dose` function does error checking/handling at each step of the procedure and
attempts to collate and report the errors in a safe manner. If you get an actual exception thrown
when using it, let me know what it is and I'll try and add it. You can see what errors were thrown
with the following:
```{r geterrors}
get_errors(mod)
```
This should return `FALSE` if no errors were thrown.

You can see diagnostics and approximation quantities too:
```{r getapprox}
# Diagnostics: U at estimated BMD and Psi at estimated score BMDL.
# Both should be zero:
get_uxb(mod)
get_psixl(mod)
# Approximation quantities: variance and derivative of U at BMD (used for delta method)
get_approximations(mod)
# Computation times of each step
get_computation_times(mod)
```

# New Smoothing Method using TMB

The method also is implemented from scratch using the Template Model Builder (TMB; https://kaskr.github.io/adcomp/). The `mgcv` package is still used to set up the bases and penalties, but then all the computations are done using custom code. The following methodological differences are present over `scam`:

- The use of Laplace-approximate marginal likelihood to estimate smoothing parameters instead of generalized cross-validation,
- The use of full integrated second derivative penalties for the monotone smooths (and other smooths) instead of difference-based P-spline penalties, and
- The direct application of linear sum-to-zero constraints on the monontone smooth, as opposed to constraint-absorbing reparameterizations. The latter are still used for the non-monotone smooth terms.

Simulations in paper show substantially improved performance of doing this way compared to `scam`.

Here are two examples. Currently the method supports (a) a single monotone smooth, plus (b) any number of non-monotone smooths, and (c) no linear terms (although these could probably be provided as part of the `smooths` argument, I have not tested this). 

## A single monotone dose response curve only

Here is an example of a model with only a single monotone smooth and nothing else. The interface is a bit different and is
documented. **Note**: important, you use the regular `bs='bs'` basis now for the monotone dose-response smooth,
**NOT** the `scam`-type `bs='mpd'`. The monotonicity is handled internally and is controlled by you passing
this formula to the `monosmooths` argument of `benchmark_dose_tmb`.
```{r monosim}
set_parameters_mono <- function(scale=1,sigma=1,p0=GLOBAL_P0,BMR=.025,plot_it=FALSE,knots=10) {
  xmin <- 0
  xmax <- 1
  # True function f
  stopifnot(scale>0)
  f <- function(x) exp(-x*scale)
  finv <- function(x) -log(x)/scale

  x0 <- 0
  stopifnot(x0>=xmin)
  tau0 <- f(x0) + sigma * qnorm(p0) # Response level
  stopifnot(abs(pnorm(tau0,f(x0),sigma) - p0) < sqrt(.Machine$double.eps)) # Definition of tau

  A <- qnorm(p0+BMR) - qnorm(p0)
  # Benchmark dose
  xb <- finv(f(x0) - sigma*A)
  stopifnot(abs(pnorm(tau0,f(xb),sigma) - (p0+BMR)) < sqrt(.Machine$double.eps)) # Definition of benchmark dose

  if (plot_it) {
    curve(f,xmin,xmax,ylim=c(0,1))
    abline(v=xb,lty='dashed')
    cat("xb = ",xb,"; f'(xb)/sigma = ",numDeriv::grad(f,xb)/sigma,".\n",sep="")
  } else {
    list(
      xmin = xmin,xmax = xmax,
      sigma = sigma,
      p0 = p0,BMR = BMR,
      f = f,
      x0 = x0,
      tau0 = tau0,
      A = A,
      xb = xb,
      knots=knots
    )
  }
}

simulate_data_mono <- function(n,params) {
  xcov <- with(params,seq(xmin,xmax,length.out=n))
  with(params,data.frame(y = rnorm(n,f(xcov),sigma),x = xcov))
}

set.seed(472389)
library(mgcv) # Need this for s()
params2 <- set_parameters_mono(scale=1,sigma=.5,p0=.01,BMR=.01,knots=10)
dat2 <- simulate_data_mono(1000,params2)

mod1_tmb <- benchmark_dose_tmb(
  monosmooths = list(s(x,bs='bs')),
  smooths = NULL,
  linearterms = NULL,
  data = dat2,
  exposure = 'x',
  response = 'y',
  x0 = 0,
  p0 = .01,
  BMR = .01,
  verbose = FALSE,
  eps = 1e-06,
  maxitr = 10,
  bayes_boot = 1e03
)
class(mod1_tmb)
```

The `benchmark_dose_tmb` function is documented and does the work here. Because the `benchmark_dose_tmb` function 
returns an object of class `semibmd`, the same `plot`, `summary`, and other methods above work for it:

```{r monosummary}
summary(mod1_tmb)
plot(mod1_tmb)
```

The output of the plot method is expanded a bit: I include a spaghetti plot of posterior samples of the fitted curve to enhance it
a bit.

We can compare to the same model fit to the same data the old way:

```{r monosimscam}
mod1_scam <- benchmark_dose(
  y ~ s(x,bs='mpd'),
  data = dat2,
  exposure = 'x',
  x0 = 0,
  p0 = .01,
  BMR = .01,
  verbose = FALSE,
  eps = 1e-06,
  maxitr = 10
)
summary(mod1_scam)
plot(mod1_scam)
```

We can the the BMD/L are a bit different:

```{r monobmdlcompare}
get_bmd(mod1_tmb)
get_bmd(mod1_scam)
get_all_bmdl(mod1_tmb)
get_all_bmdl(mod1_scam)
```

Again, simulations show the `TMB` approach is apparently better. Note that in the `plot` method for `scam`
the curve is shown **without** the intercept, where for `TMB` it is shown **with** the intercept. This can be changed if
necessary and is pretty arbitrary. The sum of the curve at the exposure values equals $0$ in what `scam` is plotting
and equals the estimated intercept in what `TMB` is plotting; again, pretty arbitrary choice and also the BMD doesn't depend on
this.

## A monotone dose response curve and two arbitrary smooths

Here is an example with a monotone dose-response smooth and two other smooths. **Note** the use of the `bs='bs'` for the
monotone dose-response curve, **not** the `scam`-type `bs='mpd'`.

```{r multismoothsim}
simulate_data_multi <- function(n,params) {
  xcov <- with(params,seq(xmin,xmax,length.out=n))
  z1 <- with(params,runif(n,xmin,xmax))
  z2 <- with(params,runif(n,xmin,xmax))


  fz1 <- function(z) sin(2*pi*z)
  fz2 <- function(z) cos(2*pi*z)
  # fz2 <- function(z) dnorm(2*z-.5)/dnorm(0)



  with(params,data.frame(y = rnorm(n,f(xcov)+fz1(z1)+fz2(z2),sigma),x = xcov,z1=z1,z2=z2))
}


set.seed(472398)
params3 <- set_parameters_mono(scale=5,sigma=.5,p0=.01,BMR=.01,knots=10)
dat3 <- simulate_data_multi(500,params3)

mod2_tmb <- benchmark_dose_tmb(
  monosmooths = list(s(x,bs='bs')),
  smooths = list(s(z1,bs='bs'),s(z2,bs='bs')),
  linearterms = NULL,
  data = dat3,
  exposure = 'x',
  response = 'y',
  x0 = 0,
  p0 = .01,
  BMR = .01,
  verbose = TRUE,
  eps = 1e-06,
  maxitr = 10,
  bayes_boot = 1e03
)

summary(mod2_tmb)
params3$xb # True BMD
```

The `plot` method works a little bit different: since there is more than one curve 
to plot, it will use the same `Hit <Return> to see next plot` thing that other packages do.

```{r multismoothplot}
plot(mod2_tmb)
```

You will probably want finer control over the plot data, to make them look nice. You
can call `plot` with `plot=FALSE` (sorry for the stupid naming) to get the data for the plot,
and then make the plot(s) yourself so you can add titles and so on:

```{r multismoothplotmanual}
plotdata <- plot(mod2_tmb,plot=FALSE)
length(plotdata)
Map(names,plotdata)
# Plot the monontone smooth
with(plotdata[[1]],plot(x,fitted[ ,1],type='l',col=scales::alpha('lightgrey',0.2),xlim = range(x),ylim = c(min(lower),max(upper)),main = "My custom plot title",xlab = "Exposure",ylab = "Response"))
M <- 100 # Or whatever
for (i in 2:M)
  with(plotdata[[1]],graphics::lines(x,fitted[ ,i],col=scales::alpha('lightgrey',0.2)))

with(plotdata[[1]],graphics::lines(x,estimate))
with(plotdata[[1]],graphics::lines(x,lower,lty='dashed'))
with(plotdata[[1]],graphics::lines(x,upper,lty='dashed'))
with(plotdata[[1]],graphics::abline(v = get_bmd(mod2_tmb)[1],lty='longdash'))
with(plotdata[[1]],graphics::abline(v = get_bmd(mod2_tmb)[2],lty='dotdash'))
```

The plot data for the other smooths is identical.

